---
title: "Seattle Crime"
author: "Alexander Valentino Evans (ave5@uw.edu), Saranya Gores (saranya@uw.edu), Noah Adam Hellyer (hellynoa@uw.edu), Ryan Trieu (dtrieu99@uw.edu)"
date: "2/18/2022"
output: html_document
---
**Affiliation**: "INFO-201: Technical Foundations of Informatics - The Information School - University of Washington"

**Abstract** :  We are interested in identifying trends within crime rates in Seattle. This is an important topic to understand because it can ultimately be used to influence policy that makes our lives safer. To address this question, we will utilize data collected over the past decade regarding crime rates in Seattle.  

**Keywords** :  Seattle crime, Violent Crime, Property Crime, Income by Location, Crime rate and Income


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
source("../source/summaryscript.R")
```
## Seattle Crime Report
# 1.0 Introduction 
Crime is an issue that many people deeply care about. Crime influences important issues such as our economy and political opinions. Crime rates can vary wildly between different neighborhoods within a city. As an example, downtown Seattle is known for having a higher rate of crime compared to the rest of the city. Our area of research for this project intends to dive deeply into crime data and understand trends that exist within that data. Additionally, we want to identify what kinds of correlations we can draw with crime from other data such as income rates. We will utilize two datasets containing crime and income data within Seattle neighborhoods over the past decade. Our goal is to capture meaningful information that Seattle politicians can utilize to make informed policy decisions that reduce crime in our city. We all would like to make Seattle a safer place to live.

# 2.0 Design Situation 

# 3.0 Research Questions

# 4.0 The Dataset

# 5.0 Expected Implications

# 6.0 Limitations

# Acknowledgements
We have no acknowledgements at this time.

## References

[^1]: [Halpern, D. (2001). MORAL VALUES, SOCIAL TRUST AND INEQUALITY: Can Values Explain Crime? The British Journal of Criminology, 41(2), 236–251.]( http://www.jstor.org/stable/23638820) .

[^2]: [The benefits of criminal justice data: Beyond policing](https://sunlightfoundation.com/2015/05/01/the-benefits-of-criminal-justice-data-beyond-policing/) .

[^3]: [Identifying Criminal Justice Stakeholders](https://study.com/academy/lesson/identifying-criminal-justice-stakeholders.html#:~:text=Internal%20stakeholders%20include%20police%20officers,court%20personnel%2C%20and%20other%20individuals.) .

[^4]: [Prieto Curiel, R., Cresci, S., Muntean, C.I. et al. Crime and its fear in social media. Palgrave Commun 6, 57 (2020). ](https://doi.org/10.1057/s41599-020-0430-7)

Here are the top 10 crime parent groups 
```{Crime parent group top 10}
kable(crime_data_top_10, col.names = c("Country", "Change in Life Expectancy"))
```
## Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


